### 2017-02-28 2017mergingmerges/section/03-Definition.tex
Shift discussion of fusion order to Fusion In Practice section, and elaborate
notes
Editing S3.1

A $\Proc$ is a record with five fields: the @ins@ field specifies the input channels; the @outs@ field the output channels; the @heap@ field the process-local heap; the @label@ field the label of the currently being executed instruction, and the @instrs@ a map of labels to instructions. We use the same record when specifying both the definition of a particular process, as well as when giving the evaluation semantics. When specifying a process the @label@ field gives the entry-point to the process code, though during evaluation it is the label of the instruction currently being executed. Likewise, when specifying a process we usually only list channel names in the @ins@ field, though during evaluation they are also paired with their current $\InputState$. If an $\InputState$ is not specified we assume it is `none'.

In the grammar of Fig.~\ref{fig:Process:Def} the $\InputState$ has three options: @none@, which means no value is currently stored in the associated stream buffer variable, $(@pending@~\Value)$ which gives the current value in the stream buffer variable and indicates that it has not yet been copied into a process-local variable, and @have@ which means the pending value has been copied into a process-local variable. The $\Value$ attached to the @pending@ state is used when specifying the evaluation semantics of processes. During fusion, the $\Value$ itself will not be known, but we can still reason statically that a process must be in the @pending@ state. We will use a different version of $\InputState$ in \S\ref{s:Fusion} when we define the fusion algorithm.

The @instrs@ field of the $\Proc$ maps labels to instructions. The possible instructions are: @pull@, which pulls the next value from a channel into a given heap variable; @push@, which pushes the value of an expression to an output channel; @case@ which branches depending on the result of a boolean expression; @jump@ which causes control to move to a new instruction, and @drop@ which indicates that the current value pulled from a channel is no longer needed.

@@

As with Kahn processes~\cite{kahn1976coroutines}, pulling from a channel is blocking. Unlike Kahn processes, pushing to a channel can also block. Each consumer has a single element buffer, which is the stream buffer variable, and pushing can only succeed when that buffer is empty.

When lowering process code to a target language, such as C, LLVM, or some sort of assembly code, we can safely ignore the @drop@ instructions. The @drop@ instructions are used to help control how processes being fused should be synchronized, but do not affect the execution of a single process. We will discuss @drop@s further in \S\ref{s:Optimisation}.

@@

Execution of process networks consists of three aspects:

\begin{enumerate}
\item \emph{Injection}, which attempts to provide a single value gained from a stream to a process, or a set of processes. Each individual process only needs to accept the value when it is ready for it, and injection of a value into a set of processes succeeds only when they \emph{all} accept it.

\item \emph{Shaking}, which advances a single process from one instruction to another. Shaking a set of processes succeeds when \emph{any} of the processes in the set can advance.

\item \emph{Feeding}, which manages communication between separate processes in the network. Feeding alternates between Injecting and Shaking. When a process pushes a value to an output channel we attempt to inject this value into all processes that have that same channel as an input. If they all accept it then we then advance their programs as far as they will go, which may cause more values to be pushed to output channels, and so on.

\end{enumerate}

Evaluation of a process network is non-deterministic. At any moment several processes may be able to take a step, while others are blocked trying to pull or push to their channels. However, because each process itself is deterministic, and uses blocking reads, the sequence of values pushed to each stream is deterministic, as per Kahn process networks.

Importantly, it is the order in which values are \emph{pushed to each particular output channel} which is deterministic, whereas the order in which different processes execute their instructions, is not. When we fuse two processes together we exploit this fact by choosing one particular instruction ordering that enables the process network to advance without requiring unbounded buffering.

Each output channel may be pushed to by a single process only, so in a sense each output channel is ``owned'' by a single process. The only intra-process communication is via channels and streams. Our model is ``pure data flow'' (or perhaps ``functional data flow'') as there are no side-channels between the processes. This is in contrast to systems such as StreamIt~\cite{thies2002streamit}, where the processes are also able to send asynchronous messages to each other, in addition to the formal input and output streams.

% -----------------------------------------------------------------------------
\subsubsection{Injection}
Fig.~\ref{fig:Process:Eval:Inject} gives the rules for injecting values into processes. The statement $(\ProcInject{p}{v}{c}{p'})$ reads ``given process $p$, injecting value $v$ into channel $c$ yields a new process $p'$''. The @injects@ form is similar, but operates on sets of processes rather than a single one.

Rule (InjectValue) injects of a single value into a single process. The value is stored as a (@pending@~ v) binding in the $\InputState$ for the associated channel of the process. The $\InputState$ acts as a single element buffer, and must be empty (set to @none@) for the injection to succeed.

@@

Rule (InjectMany) attempts to inject a single value into a set of processes. We use the single process judgment form to inject the value into all processes in the set, which must succeed for all of the. Once a value has been injected into all consuming processes that require it, it is then safe for a producing process to discard it.

@@

% -----------------------------------------------------------------------------
\subsubsection{Shaking}
Fig.~\ref{fig:Process:Eval:Shake} gives the rules for advancing a single process. The first set of rules handle specific instructions. The statement $(\ProcBlockShake{i}{is}{\Sigma}{\alpha}{l}{is'}{us})$ reads ``instruction $i$, given channel states $is$ and the heap $\Sigma$, passes control to instruction at label $l$ and yields new channel states $is$, heap update expressions $us'$, and performs an output action $\alpha$.'' An output action $\alpha$ is a list of messages of the form $(\Push~\Chan~\Value)$, which encode the values a process pushes to its output channels. We write ~$\cdot$~ for an empty action.

@@

\eject{}
The statement $\ProcShake{ps}{\alpha}{ps'}$ reads ``the processes group $ps$ advances to the new process group $ps'$ yielding output action $\alpha$. A process ``group'' is just a set of processes.

@@

% -----------------------------------------------------------------------------
\subsection{Non-deterministic Evaluation Order}

>>>>

A $\Proc$ is a record with five fields: the @ins@ field specifies the input channels; the @outs@ field the output channels; the @heap@ field the process-local heap; the @label@ field the label of the instruction currently being executed, and the @instrs@ a map of labels to instructions. We use the same record when specifying both the definition of a particular process, as well as when giving the evaluation semantics. When specifying a process the @label@ field gives the entry-point to the process code, though during evaluation it is the label of the instruction currently being executed. Likewise, when specifying a process we usually only list channel names in the @ins@ field, though during evaluation they are also paired with their current $\InputState$. If an $\InputState$ is not specified we assume it is `none'.

In the grammar of Fig.~\ref{fig:Process:Def} the $\InputState$ has three options: @none@, which means no value is currently stored in the associated stream buffer variable, $(@pending@~\Value)$ which gives the current value in the stream buffer variable and indicates that it has not yet been copied into a process-local variable, and @have@ which means the pending value has been copied into a process-local variable. The $\Value$ attached to the @pending@ state is used when specifying the evaluation semantics of processes. When performing the fusion transform the $\Value$ itself will not be known, but we can still reason statically that a process must be in the @pending@ state. When defining the fusion transform in \S\ref{s:Fusion} we will used a version of $\InputState$ with only this staticaly known information.

The @instrs@ field of the $\Proc$ maps labels to instructions. The possible instructions are: @pull@, which pulls the next value from a channel into a given heap variable; @push@, which pushes the value of an expression to an output channel;  @drop@ which indicates that the current value pulled from a channel is no longer needed; @case@ which branches based on the result of a boolean expression, and @jump@ which causes control to move to a new instruction.

@@

\eject{}
When lowering process code to a target language, such as C, LLVM, or some sort of assembly code, we can safely convert @drop@ to plain @jump@ instructions. The @drop@s instructions are used to control how processes should be synchronized, but do not affect the execution of a single process. We will discuss @drop@s further in \S\ref{s:Optimisation}.

% This allows us to \emph{deliberately} introduce artificial deadlocks when a process network would require more than one element of buffering.
%%% AR: added to highlight that this rules out networks that require unbounded buffers
%%% BL: We don't have any examples of explicitly introducing deadlocks. The process networks just happen to have them when viewed abstractly.

%%% AR: feels a bit disjointed because drops were only mentioned once a few paragraphs ago. Maybe reword to talk about lowering in general is obvious for most instructions, and drops are just treated as jumps. Or move up.

@@

The dynamic execution of process networks consists of three aspects:

\begin{enumerate}
\item \emph{Injection} of a single value from a stream into a process, or a set of processes. Each individual process only needs to accept the value when it is ready for it, and injection of a value into a set of processes succeeds only when they \emph{all} accept it.

\item \emph{Advancing} a single process from one state to another. Advancing a set of processes suceeds when \emph{any} of the processes in the set can advance.

\item \emph{Feeding} outputs of some processes to the inputs of others. Feeding alternates between Injecting and Advancing. When a process pushes a value to an output channel we attempt to inject this value into all processes that have that same channel as an input. If they all accept it then we then advance their programs as far as they will go, which may cause more values to be pushed to output channels, and so on.
\end{enumerate}

Execution of a process network is non-deterministic. At any moment several processes may be able to take a step, while others are blocked. As with Kahn processes~\cite{kahn1976coroutines}, pulling from a channel is blocking. This enables the overall sequence of values on each output channel to be deterministic. Unlike Kahn processes, pushing to a channel can also block. Each consumer has a single element buffer, which is the stream buffer variable, and pushing can only succeed when that buffer is empty.

%%% AR: what is the distinction between 'execution' and 'evaluation'?  I only have a vague feeling that execution is something a computer does, while evaluation is the mathematical rules. Either way, these should probably be consistent.
%%% BL: "Evaluation" is pure.  E-"value"-ation. Execution has visible actions, like pushing to streams.

Importantly, it is the order in which values are \emph{pushed to each particular output channel} which is deterministic, whereas the order in which different processes execute their instructions, is not. When we fuse two processes we exploit this fact by choosing one particular instruction ordering that enables the process network to advance without requiring unbounded buffering.

Each output channel may be pushed to by a single process only, so in a sense each output channel is owned by a single process. The only intra-process communication is via channels and streams. Our model is ``pure data flow'' (or perhaps ``functional data flow'') as there are no side-channels between the processes. This is in contrast to systems such as StreamIt~\cite{thies2002streamit}, where the processes are also able to send asynchronous messages to each other, in addition to via the formal input and output streams.

% -----------------------------------------------------------------------------
\subsubsection{Injection}
Fig.~\ref{fig:Process:Eval:Inject} gives the rules for injecting values into processes. The statement $(\ProcInject{p}{v}{c}{p'})$ reads ``given process $p$, injecting value $v$ into channel $c$ yields an updated process $p'$''. The @injects@ form is similar, but operates on sets of processes rather than a single one.

Rule (InjectValue) injects of a single value into a single process. The value is stored as a (@pending@~ v) binding in the $\InputState$ of the associated channel of the process. The $\InputState$ acts as a single element buffer, and must be empty (set to @none@) for the injection to succeed.

@@

Rule (InjectMany) attempts to inject a single value into a set of processes. We use the single process judgment form to inject the value into all processes in the set, which must succeed for all of them. Once a value has been injected into all consuming processes that require it, the producing process no longer needs to retain it.

@@

\eject{}
% -----------------------------------------------------------------------------
\subsubsection{Advancing}
Fig.~\ref{fig:Process:Eval:Shake} gives the rules for advancing a single process. The first set of rules handle specific instructions. The statement $(\ProcBlockShake{i}{is}{bs}{a}{l}{is'}{us'})$ reads ``instruction $i$, given channel states $is$ and the heap bindings $bs$, passes control to instruction at label $l$ and yields new channel states $is'$, heap update expressions $us'$, and performs an output action $a$.'' An output action $a$ is an optional message of the form $(\Push~\Chan~\Value)$, which encodes the value a process pushes to one of its output channels. We write ~$\cdot$~ for an empty action.

@@

\eject{}
The statement $\ProcShake{ps}{\alpha}{ps'}$ reads ``the processes group $ps$ advances to the new process group $ps'$ yielding output action $\alpha$. A process ``group'' is just a set of processes. 
%%% AR: why `group' and not `network'? I think network, as in Kahn process network, is fairly standard terminology

@@

% -----------------------------------------------------------------------------
\subsection{Non-deterministic Evaluation Order}
\label{s:EvaluationOrder}
%%% AR: Can we simplify this?
% Maybe we can start by talking about the different evaluation orders for just two alt2s:
%
% let s1 = alt2 sInA sInB
% let s2 = alt2 sInA sInC
%
% Here we have many evaluation orders. If we just look at the order in which inputs sInA etc are pulled, some options are:
% [ A, A, B, C, B, C ]
% [ A, A, B, B, C, C ]
% [ A, A, C, C, B, B ]
%
% Now if we add a new operator which zips the two together, it actually restricts the pulling order
%
% let s1 = alt2 sInA sInB
% let s2 = alt2 sInA sInC
% let s3 = zip  s1 s2
%
% This restricts the pulling order to:
% [ A, A, B, C, B, C ]
%
%%%
% Or, maybe we don't even need the zip here, and can leave the zip until S5.1.
%%%
